\Large Last Writer Slices and Communication Traps

### 2013-11-06 paper.tex
committing this little daddy after a onceover and a new conclusion.

\section{System Implementation and Applications}

@@

We released our implementation, free and open-sourced~\cite{ctrapsonweb}, on
the web and have had contributions to our codebase from members of the GCC
development community.

@@

%\begin{table}
%\begin{tabular}{ l | l | c | l }
%& & & \\ \hline
%\multirow{9}{*}{\begin{sideways}PARSEC\end{sideways}} & blackscholes & & \\
%&dedup & & \\
%&canneal & & \\
%&x264& & \\
%&vips& & \\
%&ferret& & \\
%&fluidanimate& & \\
%&swaptions & & \\
%&streamcluster & & \\ \hline
%\multirow{4}{*}{\begin{sideways}Servers\end{sideways}}&MySQL & & \\
%&Apache & & \\
%&LevelDB & & \\
%&Memcached & & \\ \hline
%\multirow{3}{*}{\begin{sideways}App\end{sideways}}&PBZip2 & & \\
%&AGet & & \\
%&Transmission & & \\
%
%\end{tabular}
%\caption{\label{tab:benchmarks}\addtodo{build this table, show LOC, \#threads, sharing pattern(?), etc}}
%\end{table}

@@

The main performance result of our work is that our designs impose performance
overheads that are low enough for use in deployed systems for many of our
experimental workloads.  Figure~\ref{fig:perfall} shows the slowdown suffered
by our benchmarks due to last writer slicing and \ctraps running with ``no-op''
trap hooks ({\em i.e.}, no handlers), relative to the natively executing
baseline.  These data reflect our main performance findings.

\paragraph{Last Writer Slices}
The data show that last writer slicing has very low overheads with a geometric
mean of less than 10\% across our server programs and less than 50\% across
PARSEC.  Such low overheads are very likely to be tolerable in deployed
systems.  In many cases (Apache, MySQL, {\tt dedup}, {\tt canneal}), overheads
are negligible.  In all but two cases ({\tt vips} \& {\tt swaptions}), the
overhead of collecting last writer slices is less than 100\%.

@@

By comparing \ctraps and last writer slicing, we see four applications ({\tt
dedup}, {\tt streamcluster}, {\tt fluidanimate}, {\tt ferret}) that have
\ctraps overhead that is probably too high for always-on use (averaging 153\%)
drop to a last writer slicing overhead that is probably low enough for
always-on use (averaging 38\%).  The large decrease indicates these programs
perform relatively more reads than other applications, and thus benefit more
from eliminating read instrumentation.

@@

\subsubsection{Overheads Due to Conservative Escape Analysis}
\label{sec:eval:conservative}
{\tt swaptions} had the highest run time overhead in our tests.  We
investigated the cause and determined it is largely due to shortcomings in our
escape analysis.  We located the inner-loop function {\tt
HJM\_SimPath\_Forward\_Blocking}.  The function uses two local matrices that
are allocated through an external call.  Our escape analysis cannot determine
that the call only does allocation and that the matrices are local, so it
conservatively assumes the matrices escape.  We manually inlined the allocation
(and a matching deallocation), and escape analysis eliminated their
instrumentation.  The program's overhead dropped from about 420\% overhead for
\ctraps to about 300\% overhead; the change for last writer slicing was from
300\% to about 130\% overhead.  We also looked at an output matrix passed into
the inner-loop function.  The matrix is allocated and deallocated in the
inner-loop's caller, but is never shared.  Eliminating instrumentation on
accesses to this matrix yielded an overhead of about 46\% for \ctraps, and to
just 30\% for last writer slicing.

Note that both these changes amount to inlining and preserve the semantics and
sharing pattern of the program.  We expect a more sophisticated escape
analysis would find these local accesses and achieve these performance gains
automatically.

@@

\subsubsection{Performance with \ctraps}
Figure~\ref{fig:perfapps} shows the performance overheads of our implementation
as the slowdown with respect to the natively executing baseline.  The overheads
vary substantially across applications.  On average, the overheads are higher
than for our baseline implementation (Section~\ref{sec:eval:perf}).

In six cases (all servers, {\tt blackscholes}, and {\tt dedup}), the overheads
for CCI-Prev implementations are tolerable for deployment use (around 50\%).
This result supports our claim that \ctraps helps build useful analyses that
have overheads low enough for deployment for an important class of
applications.

@@

@@

\begin{table}
%\renewcommand{\tabcolsep}{2pt}
\centering
\small
\begin{tabular}{l | r r }
\multirow{3}{*}{\bf Application} & \multicolumn{2}{c}{\bf \em Pruned } \\
               & \multicolumn{2}{c}{\bf \em Static Local}  \\ 
               & \multicolumn{2}{c}{\bf \em Operations}    \\ \hline

@@

@@

Thin Slicing~\cite{thinslicing} is related in its approach. Thin
Slicing monitors and reports a select subset of dependent operations from an
execution.  The goal is to eliminate some operations that would be included in a traditional
slice to make the slice simpler and less costly to collect.  While this work and
Thin Slicing differ in what they collect and how, our motivation to monitor a
select subset of dependences was similar to theirs.    Prior work
on slicing is unlike our work in that little has focused on monitoring
inter-thread dependences in production code without sampling or hardware support.

@@

CCI~\cite{cci}, Bugaboo~\cite{bugaboo}, Recon~\cite{recon}, and
DefUse~\cite{defuse} all explicitly track inter-thread memory dependences at
runtime.  Unlike our work, these techniques were not designed to run in
production systems~\cite{recon,defuse} without sampling~\cite{cci} or hardware
support~\cite{bugaboo}.

Some other techniques~\cite{threadclustering,threadcriticality} have relied on
hardware debugging registers to monitor dependences efficiently enough for
production use.  These techniques are similar to our work in that they track how
threads interact ({\em i.e.}, share).  These techniques are different in that
their approach to dependence tracking is specific to the problem they address;
our system provides general support.

\paragraph{Extensible Program Instrumentation}
There are many systems for building analyses using extensible program
instrumentation.  Binary instrumentation
systems~\cite{pin,dynamorio,valgrind,roadrunner} provide completely general
support to instrument arbitrary code -- instrumenting a program to track
communciation is possible in such systems.  These systems are similar to
\ctraps in that they aim to enable dynamic analyses and runtime systems.  These
are different from \ctraps because they rely on dynamic rewriting, yielding
overheads too high for deployment.  \ctraps provides extensible support for a
restricted set of instrumentation tasks ({\em i.e.}, tracking communication)
with overheads low enough for production.  Such general systems also increase
the burden on implementors: to monitor communication, they must manually track
dependences, crafting delicate, performance sensitive code.  \ctraps tracks
only those dependences corresponding to communcation.  Furthermore, 
instrumentation systems lack the debugging benefits afforded by last writer slices.

\section{Conclusions}
This work stems from the observation that monitoring inter-thread communication
is often fast enough for use in deployed systems.  Based on this premise, we
develop system support for efficiently collecting last writer slices.  Using
that support, we developed \ctraps, a system for extensibly tracking
communicating inter-thread dependences.  \ctraps works efficiently at runtime,
in production, and without sampling or hardware support.  We showed that
\ctraps exposes a sound and complete (assuming race freedom) communication
tracking interface, enabling trap handlers to implement applications in just a
few lines of code.  We showed that a useful class of analyses can be
implemented using \ctraps that are efficient enough for deployment use in an
important set of programs ({\em e.g.}, servers, key-value stores, databases,
{\em etc.}).  We showed that last writer slices provide information that is
helpful for understanding concurrency errors with nearly no overhead.  Our
future work includes studying the generality of possible \ctraps applications
and the role of minimal hardware support for \ctraps.  We will also further study the
impact of provenance information on debugging in both sequential and concurrent
applications.

>>>>

\section{System Implementation}

@@

\subsection{Release}
We released our implementation, free and open-sourced, on the
web~\cite{withheldforanonymity} and published it to the GCC plugin repository.
Since its release, we have seen contributions to our codebase from members of
the GCC development community and downloads from interested developers.

@@

@@

The main performance result of our work is that our designs impose performance
overheads that are low enough for use in deployed systems for many of our
experimental workloads.  Figure~\ref{fig:perfall} shows the slowdown suffered
by our benchmarks due to last writer slicing and \ctraps running with ``no-op''
trap hooks ({\em i.e.}, no handlers), relative to the natively executing
baseline.  In these experiments, we consider a run time overhead of 10\% or
less to be ideal for production use and a run time overhead of 50\% as
a reasonable upper bound on acceptable overhead for production use.

\paragraph{Last Writer Slices}
The data show that last writer slicing has very low overheads with a geometric
mean of less than 10\% across our server programs and less than 50\% across
PARSEC.  Such low overheads are likely to be acceptable in production.  In many
cases (Apache, MySQL, {\tt dedup}, {\tt canneal}), overheads are negligible.
In all but two cases ({\tt vips} \& {\tt swaptions}), the overhead of
collecting last writer slices is less than 100\%.

@@

By comparing \ctraps and last writer slicing, we see four applications ({\tt
dedup}, {\tt streamcluster}, {\tt fluidanimate}, {\tt ferret}) that have
\ctraps overhead that is probably too high for production use (averaging
153\%).  By comparison, those applications have a last writer slicing overhead
that is likely acceptable for production use (averaging 38\%).  The difference
indicates that these programs perform relatively more reads than other
applications, experiencing more overhead due to \ctraps's read instrumentation.

@@

\subsubsection{Overheads Due to Conservative Escape Analysis}
\label{sec:eval:conservative}
{\tt swaptions} had the highest run time overhead in our tests.  We
investigated the cause and determined it is largely due to shortcomings in our
escape analysis.  We located the inner-loop function {\tt
HJM\_SimPath\_Forward\_Blocking}.  The function uses two local matrices that
are allocated through an external call.  Our escape analysis cannot determine
that the matrices are only used locally in that function, so it conservatively
assumes the matrices escape.  We manually inlined the matrix allocation calls
(and a matching deallocation calls), and escape analysis eliminated their
instrumentation.  The program's overhead dropped from about 420\% overhead for
\ctraps to about 300\% overhead; the change for last writer slicing was from
300\% to about 130\% overhead.  We also looked at an output matrix passed into
the inner-loop function.  The matrix is allocated and deallocated in the
inner-loop's caller, but is never shared.  Eliminating instrumentation on
accesses to this matrix yielded an overhead of about 46\% for \ctraps, and to
just 30\% for last writer slicing.

These manual improvements to the escape analysis reduced the run time overheads
for these workloads from being unacceptable for any production environment to
being reasonable for use in production systems.  We expect a more sophisticated
escape analysis would find these local accesses and achieve these performance
gains automatically.

@@

\subsubsection{Performance with \ctraps}

 Figure~\ref{fig:perfapps} shows the performance overheads of
our implementation as the slowdown with respect to the natively executing
baseline.  The overheads vary substantially across applications and the average
run time overheads are higher than for our baseline implementation
(Section~\ref{sec:eval:perf}).

In six cases (the four servers, {\tt blackscholes}, and {\tt dedup}), the overheads
for CCI-Prev implementations are tolerable for deployment use (around 50\%).
This result supports our claim that \ctraps helps build useful analyses that
have overheads low enough for deployment for an important class of
applications.

@@

{\bf TODO: Compare with these analyses implemented using binary instrumentation}

@@

\begin{table}
\centering
\small
\begin{tabular}{l | r r }
\multirow{3}{*}{\bf Application} & \multicolumn{2}{c}{\bf \em Pruned } \\
               & \multicolumn{2}{c}{\bf \em Static Local}  \\ 
               & \multicolumn{2}{c}{\bf \em Operations}    \\ \hline

@@

{\bf TODO: Run without escape analysis?}

@@

Thin Slicing~\cite{thinslicing} is especially related to this work. Thin
Slicing monitors and reports a select subset of dependent operations from an
execution.  The goal is to eliminate some operations that would be included in
a traditional slice to make the slice simpler and less costly to collect.
While this work and Thin Slicing differ in what they collect and how, our
motivation to monitor a select subset of dependences was similar to theirs.

%is unlike our work in that our work focused on
%monitoring inter-thread dependences in production code without sampling or
%hardware support.

\paragraph{Debugging with Data Provenance Information}

Prior work has looked at using data provenance information in
debugging~\cite{whylineicse, whylinechi}.  The WhyLine debugger allowed
programmers to ask questions about why particular variables held particular
values during debugging.  WhyLine answered those questions by presenting data-
and control-flows that influenced those variables.  WhyLine showed that
provenance information is intuitive to programmers and makes debugging easier,
justifying our focus on provenance information in our last writer slicing
design.

There are several characteristics of this work that distinguish it from prior
work on debugging with data provenance information.  Unlike prior work, our
work focuses on isolating a limited subset of dependence information that we
can collect with run time overheads low enough for production use.
Furthermore, we focus on {\em concurrent} program debugging.  Prior work on
provenance debugging that focused mainly on single threaded programs and had
dependence slicing overheads are probably too high for production.

@@

CCI~\cite{cci}, Bugaboo~\cite{bugaboo}, Recon~\cite{recon}, and
DefUse~\cite{defuse} all explicitly track inter-thread memory dependences at
runtime.  Unlike our work, several of these techniques were not designed to run
in production systems~\cite{recon,defuse}.  A distinguishing characteristic of
our work is that it is effective and fast enough for production without the
need to aggregate information from multiple different sampled executions, like
CCI~\cite{cci}.  Additionally, our system does not require invasive hardware
changes like Bugaboo~\cite{bugaboo} or LBA~\cite{paralog}.

Other techniques~\cite{threadclustering,threadcriticality} use  hardware debug
registers to monitor dependences efficiently enough for production use.  These
techniques are similar to our work in that they track how threads interact.
These techniques are different from our work in that they track information
specific to the performance and our system provides general support.

\paragraph{Extensible Program Instrumentation}
There are many systems for building analyses using extensible program
instrumentation.  Binary instrumentation
systems~\cite{pin,dynamorio,valgrind,roadrunner} provide completely general
support to instrument arbitrary code -- instrumenting a program to track
communciation is possible in such systems.

These systems are similar to our work in that they aim to enable dynamic
analyses and runtime systems.  They are different from our work because they
rely on dynamic rewriting, yielding overheads often too high for deployment.
Instead, we trade generality for performance. Last writer slices and \ctraps
provide extensible support in the restricted domain of tracking communication
with overheads low enough for production.  Another difference is that our work
handles some of the delicate, performance sensitive code required to track
communication.  In general instrumentation systems, programmers write that code
from scratch at their own risk.

\section{Conclusions}
This work stems observes that we can collect a useful form of data provenance
information with overheads low enough for use in deployed systems.

We developed system support that collects last writer slices with overheads
well within an acceptable range for production.  We showed that, during
debugging, the provenance information provided by last writer slices reveals
crucial connections between different threads' otherwise disparate regions of
code.  Understanding how code in different threads interacts is essential to
concurrency debugging and this work helps with that understanding.

Using last writer slices, we then built \ctraps, an extensible framework for
implementing concurrent program analyses that interpose on inter-thread
communication.  Trap handlers implement such analyses in just a few lines of
code.  We showed that it is possible to build useful analyses using \ctraps
that are efficient enough for deployment use in an important set of programs
({\em e.g.}, servers, key-value stores, databases, {\em etc.}).

An interesting direction for future work is to evaluate the role of minimal
hardware support for \ctraps.  We will also further study the impact of
provenance information on debugging in both sequential and concurrent
applications.
